---
title: Using Assessment Data to Inform Instruction
author: Amrit Patel
date: '2019-06-27'
slug: using-assessment-data-to-inform-instruction
categories:
  - teaching
  - R
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
```

The following figure is a simple illustration of how students can be grouped based on abilities using [sample assessment data](https://raw.githubusercontent.com/AmritPatel/myBlog/master/content/resources/algebra1Period3.csv). Assessment scores have been averaged for each student with 1-$\sigma$ standard deviations shown as a way to measure student score consistency.

```{r, fig.width = 9, rows.print=23}
library(tidyverse)

scores <- # Format student scores for plotting
  tbl_df(read.csv("~/myBlog/content/resources/algebra1Period3.csv")) %>%
  gather(key = "assignment", value = "score", -studentID) %>%
  arrange(studentID)

scores$studentID <- as.character(scores$studentID) # studentID needs to be 'character' vs. 'integer'  for color labeling

means <- # Calculate the mean and standard deviation of the scores so we can disinguish groups
scores %>% 
  group_by(studentID) %>%
  summarise(mean = mean(score), sigma = sqrt(var(score))) %>%
  arrange(mean) 

# means %>% rmarkdown::paged_table() # This command gives us a pretty table when rendered in HTML

library(Hmisc)

ggplot(means, aes(studentID, mean, color=cut2(mean, g=4))) + # cut2 used to group 'mean' data into quantiles
  geom_point() + # Scatterplot used so standard deviation can be plotted with 'mean' data
  geom_errorbar(aes(ymin=mean-sigma, ymax=mean+sigma), width=.2, position=position_dodge(0.05)) +
  theme(legend.position = "bottom") # move the legend to the bottom to allow for wider x-axis
```
***Figure 1. Student grouping based on performance.***

Paired with the figure above, the figure below can be used to quickly identify trends for a given student by looking at performance on various assignment categories (i.e., homework, projects, quizzes, reviews, and tests). Comparing scores on homework and projects versus performance on summative assessments can be a useful indicator of student engagement or determination. For example, Student 2271 performed well on the test and test review, but didn't do as well on homework and the project -- this student may be bored or disengaged. Another example is Student 2562, who performed well on homework and the project, but not as well on the test and test review -- this student might be determined to do well by completing individual assignments, but may be struggling to tie these individual concepts together.

```{r, fig.height=10}
#ggplot(scores, aes(score, fill=studentID)) + geom_histogram()
#ggplot(scores, aes(score, fill=studentID)) + geom_histogram() + facet_grid(assignment ~ .)

# Plotting data by assignment for each student to see more detail
ggplot(scores, aes(score, fill=assignment)) + geom_histogram() + facet_grid(studentID ~ .)
```
***Figure 2. Individual student performance by assignment type.***

## Creating Mixed Student Groups of Varying Abilities and Talents for a Particular Lesson

For a future lesson involving groups, I would use the data from Figure 1 to create groups with a variety of abilities allowing stronger students to help out students not performing as well or who are disengaged. I would do this by having an approximatley equal number of students from each of the 4 quantiles,[^1] indicated in Figure 1, in a group. Increasing group diversity may also have a positive impact on idea creation by avoiding "groupthink", which is a positive attribute to this strategy.

[^1]: Quantiles conveniently breakup a large group into smaller "similar" groups assigning equal numbers to each of those smaller groups.

```{r}
library("htmltools")
library("vembedr")
embed_url("https://youtu.be/2p-uY_i-wvc")
```

## Examples of Computer Programming Activities 

Learning computer programming can be daunting for beginners as there are many rules and conventions to learn. However, there are useful strategies to implement for all types of learners!

### For Remediation

The [`swirl`](https://swirlstats.com/)[^2] package's power comes from its ability allow individual students to practice programming and/or data science *interactively* and *at one's own pace*.

[^2]: For the [R programming language](https://www.r-project.org/about.html) -- this is the language that I will teach in my class.

### For Special Needs Students

[Scratch](https://scratch.mit.edu/about/) from MIT Media Lab for students with [ADHD](https://en.wikipedia.org/wiki/Attention_deficit_hyperactivity_disorder) -- a stimulating and immersive way to teach programming.

### Acceleration Activities for Gifted/Talented Students

Use programming skills to design an [interactive visualization](https://www.r-graph-gallery.com/interactive-charts/) to answer a question, using a real-world dataset, that is important to the student.