---
title: Ma Nepali bolna sikchu.
author: Amrit D. Patel
date: '2020-10-09'
slug: ma-nepali-bolna-sikchu
categories:
  - learning
tags:
  - nepal
draft: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F, fig.align = "center")
```

## Inspiration

- A simple Nepali sentence constructor inspired by [Nepali Ann's Itty-Bitty (but darn good) Nepali Language Primer](http://the-voyagers.tripod.com/language.htm)

- [Nepalgo](https://nepalgo.de/)

## Verbs

### Notes

- The order of sentences is reversed in Nepali. The conjugated verb is **always** last.
  + Sentence structure is: *subject* + *object* + *verb*.

### Vocabulary

```{r}
library(tidyverse)
library(googlesheets4)

# Load data from GSheets spreadsheet
gs4_deauth()
data_loc <- "https://docs.google.com/spreadsheets/d/1Iw_fyccbC4IBifAOG4UfdqZ23s_5QQgn3JY7WxxV06A"
read_gsheet <- function(data_location, sheet_id){
      googlesheets4::read_sheet(data_location, sheet = sheet_id, skip = 0, col_types = "c")
}
```

```{r, fig.align = 'center'}
library(kableExtra)

# Present table of basic verbs
verbs <- read_gsheet(data_loc, "verbs")

verbs %>%
  kable() %>%
  kable_styling(full_width = F, position = "center")
```

### Conjugation

```{r}
conjugation <- read_gsheet(data_loc, "conjugation")
conjugation %>%
  kable() %>%
  kable_styling(full_width = F, position = "center")
```

## Nouns

### Notes

- There is a strong tendency for the indirect object to be placed **before** the direct object.
- If there is no indirect object and a verb is applied to an **animate** direct object, `-lai` is added. This is the *accusative* case.
- The structure which employs the dative case, which is marked by `-lai` in Nepali, is called *dative construction*, and is always applied to the indirect object of the sentence.
  + It is used mostly to describe **feelings** and **emotions**.
- To make a noun (or pronoun) plural, use pluralizer `-haru`.
  + If noun ends in `-o`, drop it, and add `-a`, then add `-haru`.
    
```{r}
nouns <- read_gsheet(data_loc, "nouns")
nouns %>%
  kable() %>%
  kable_styling(full_width = F, position = "center")
```

## Adjectives

### Notes

- If adjective ends in `-o`, drop it, and add `-a`, but don't add `-haru`.

```{r}
adjectives <- read_gsheet(data_loc, "adjectives")
adjectives %>%
  kable() %>%
  kable_styling(full_width = F, position = "center")
```

## Adverbs

- Adverbs are not declined for number.

```{r}
adverbs <- read_gsheet(data_loc, "adverbs")
adverbs %>%
  kable() %>%
  kable_styling(full_width = F, position = "center")
```

## Conjunctions

### Notes

- Conjunctions are not declined for number.

```{r}
conjunctions <- read_gsheet(data_loc, "conjunctions")
conjunctions %>%
  kable() %>%
  kable_styling(full_width = F, position = "center")
```

## Flashcard generator

```{r}
library(glue)

genSentenceNepali <- function() {
  sentenceNepali <-
  bind_cols(adjectives %>% select(adj...1 = nepali) %>% slice_sample(),
          nouns %>% select(noun...2 = nepali) %>% slice_sample(),
          adjectives %>% select(adj...3 = nepali) %>% slice_sample(),
          nouns %>% select(noun...4 = nepali) %>% slice_sample(),
          verbs %>% select(verb = nepaliRoot) %>% slice_sample()
          )
}

# flashcard generator
flashcard <- genSentenceNepali()

# flashcard wrapper
constructPresentFormal <- function() {
  flashcard %>% mutate(sentence = glue("{adj...1} {noun...2}lai {adj...3} {noun...4} {verb}chhu.")) 
}

constructPresentFormal()$sentence
```

