---
title: 30 day rope skip challenge.
author: Amrit D. Patel
date: '2020-08-08'
slug: learning-to-rope-skip
categories:
  - exploratory
tags:
  - regression
  - plot
---

Over our recent 5 month stay in Thailand we happened to acquire a jump rope. For most of the time it collected dust. Now that I've picked it up again I'm thinking the reason was because jumping rope is actually pretty challenging. On multiple levels.

There is the whole timing the rope turns in sync with your jumping, which stimulates brain activity and as [crossrope.com](https://www.crossrope.com/blogs/blog/benefits-of-jumping-the-unique-value-of-jump-rope-training-good-cardio/) explains:

> *...helps with the development of the left and right hemispheres of your brain, which further enhances spatial awareness, improves reading skills, increases memory and makes you more mentally alert.*

> *As described in [an] [article from BrainHQ](https://www.brainhq.com/brain-resources/everyday-brain-fitness/physical-exercise), the best exercise for the brain involves a combination of timing, rhythm, coordination, and mental strategy.*

> *When you’re jumping rope, you’re combining timing and rhythm while having to make quick calculations to ensure your brain is recognizing the rope’s distance, speed, direction, and position relative to your body to ensure the rotations keep going. This combination of physical and mental activity has a greater effect on cognitive function.*

Other touted benefits include:

- high efficiency calorie burning
- improved cardio
- increased muscle engagement
- improved bone density

As the [*Jump Rope Institute*](http://www.jumpropeinstitute.com/benefits.htm) says: 

> The benefits of this simple but yet complex training technique ranks as an ideal brain exercise, bone builder, and as one of the most efficient ways of improving cardiovascular fitness ***in as little as ten minutes***.

More impressively, the *Jump Rope Institute* also says that with only 10 minutes of nonstop jumping at 120 jumps per minute (jpm) it can provide the same benefits as:

- 30 minutes of jogging (**!**)
- 2 sets of tennis singles
- 30 minutes of racket/handball playing
- 720 yards of swimming
- 18 holes of golf

What a missed opportunity for some quality fitness training (if true)!

## The gear

After the first day of jumping, counting jumps wasn't my favorite part. I started a spreadsheet recording the number of jumps and my max heart rate, and quickly realized that there is likely an even better way to track this data.

I headed over to the  *Garmin Connect IQ Store* and was pleasantly surprised to find the really nice [*Jumps*](https://apps.garmin.com/en-US/apps/539e6c9e-a735-45c6-b390-c0bc65c1d65a) data field that not only counts your total jumps for you, but also gives you jumps over time data, jpm over time data, and a "Jumping Effect" score, which is supposed to be an indicator for how hard you trained with a value of 100 being equivalent to jumping 20 minutes non-stop at a 120+ jpm pace.

I also learned a new unit to describe efficiency of physical exertion: the metabolic equivalent or MET. A MET is defined as *the objective measure of the ratio of the rate at which a person expends energy, relative to the mass of that person, while performing some specific physical activity compared to a reference, set by convention at 3.5 mL of oxygen per kilogram per minute, which is roughly equivalent to the energy expended when sitting quietly*.[^1]

So, you spend 1 MET sitting around playing video games, for example, and a whopping [12.3 METs](https://sites.google.com/site/compendiumofphysicalactivities/Activity-Categories/sports) jumping rope at a pace of 120-160 jpm![^2]

[^1]: From [Wikipedia: Metabolic equivalent of task](https://en.wikipedia.org/wiki/Metabolic_equivalent_of_task).
[^2]: 100 jpm burns 11.8 METs and 80 jpm burns 8.8 METs.

## The setup

Building off of the heat training running posts from this past summer and the utilities built, I realized it would be straightforward to create another automated data pipeline. The basic flow is:

1. Pull raw data from *GarminDb* `bash` script to get *Garmin* .FIT files containing jump rope activity data from the cloud.
2. Run [`fitparse`](https://github.com/dtcooper/python-fitparse) Python script in R via the `reticulate` library to convert binary .FIT to readable ASCII.
3. Post-process ASCII file using `readr::read_csv` to get data into a dataframe, then formatting the data for plotting.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F, fig.align = "center")
library(reticulate)
virtualenv_create("jumps")
py_install("fitparse", envname = "jumps")
use_virtualenv("jumps")
```

```{r, eval = F}
# py_install("python-fitparse")
# py_config() 
# py_available(initialize = FALSE)
# py_module_available("fitparse")
# conda_list()
# use_condaenv(condaenv = "auto")
```

### Step 1: Get raw data 

```{r, echo = T}
library(DBI)
library(RSQLite) 
library(tidyverse)

# Setup a working directory to copy files to for processing.
workDir <- "~/HealthData/FitFiles/working/"
# setwd(workDir)

con <- dbConnect(SQLite(), "/Users/amrit/HealthData/DBs/garmin_activities.db")

# dbListTables(con)

acts <- dbReadTable(con, 'activities')

# Path of the SQLite activities database.
path <- "/Users/amrit/HealthData/FitFiles/Activities/"

# Get .fit filenames based on running activity names.
cardio <-
  acts %>%
  filter(sub_sport == "cardio_training") %>%
  arrange(desc(start_time)) %>%
  # Allow partial activity date matching.
  filter(str_detect(start_time, "2020-")) %>%
  select(start_time, activity_id) %>%
  mutate(activity_id = paste0(path, activity_id, ".fit"))
 
save(cardio, file = "~/HealthData/FitFiles/working/cardio.Rdata")

# Copy .fit files to working directory.
walk(paste("cp", cardio$activity_id, workDir), system) # same as lapply w/ no output printed
```

### Step 2: Convert data from binary to human readable form 

```{r, echo = T}
getFitOut <- function() {

library(reticulate)
py_capture_output(py_run_string("import fitparse

# Load the FIT file
fitfile = fitparse.FitFile(r.fitFile)

# Iterate over all messages of type \"record\"
# (other types include \"device_info\", \"file_creator\", \"event\", etc)
for record in fitfile.get_messages(\"session\"):

    # Records can contain multiple pieces of data (ex: timestamp, latitude, longitude, etc)
    for data in record:

        # Print the name and value of the data (and the units if it has any)
        if data.units:
            print(\"{}, {}, {}\".format(data.name, data.value, data.units))
        else:
            print(\"{}, {}, NA\".format(data.name, data.value))
"))
}
```

### Step 3: Format data for plotting 

```{r, echo = T}
getJumpData <- function(jumpFit) {

# Run python file created by @mcandocia on GitHub @ https://github.com/mcandocia/examples/tree/master/convert_fit_to_csv 
# to convert .fit data to .csv to allow for processing in R.
  
# source_python("jumps.py", envir = NULL)

jump <- read_csv(jumpFit, col_name = c('type', 'value', 'units'))

jump %>%
  filter(
    type == "unknown_110" |
    type == "Jumping Effect" |
    type == "Jumps" |
    type == "avg_heart_rate" |
    type == "max_heart_rate" |
    type == "avg_temperature" |
    type == "max_temperature" |
    type == "start_time" |
    type == "sub_sport" |
    type == "total_anaerobic_training_effect" |
    type == "total_training_effect" |
    type == "total_calories" |
    type == "total_elapsed_time"
    ) %>%
  select(type, value) %>%
  pivot_wider(names_from = type) %>%
  type_convert(col_types = cols(.default = col_double(),
                                start_time = col_datetime(),
                                sub_sport = col_character(),
                                unknown_110 = col_character()
                                )
               ) %>%
  rename(name = unknown_110,
         jumpingEffect = "Jumping Effect",
         jumps = Jumps
         )
}
```

```{r, echo = T}
passFit <- function() {
  getJumpData(getFitOut())
}

datalist = list()

for (i in 1:length(cardio$activity_id)) {
  fitFile <- cardio$activity_id[i]
  dat <- passFit()
  datalist[[i]] <- dat
}

jumpData <-
  bind_rows(datalist) %>%
  filter(jumps > 0) %>%
  mutate(jpm = jumps / total_elapsed_time * 60)
```

```{r, echo = T}
pltJumpData <- 
  jumpData %>%
  select(-sub_sport, -name) %>%
  pivot_longer(-start_time)

ggplot(pltJumpData, aes(start_time, value)) + 
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  facet_wrap(~ name, scales = "free_y") +
  labs(title = "Summary of jump rope session data",
       x = "Date",
       y = "Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r, eval = F}
# ggplot(jumpData, aes(start_time, jpm, color = Hmisc::cut2(jumps, cuts = c(501)))) + 
#   geom_point() +
#   geom_smooth(method = lm, se = FALSE)

# ggplot(jumpData %>% filter(jumps >= 500), aes(jpm, avg_heart_rate, label = jumps)) +
#   geom_point() +
#   geom_text(vjust = 0, nudge_y = 0.3)
```

## The challenge

The natural challenge that I gave myself based on my reading of the above info sources was to jump rope for 30 days straight. *And*, each session should last for at least ~10 minutes (or at least ~1000 jumps).

Below is the plot that I've designed to help me keep track of my daily jump data, while also allowing me to track changes in fitness by observing changes in average heart rate during each jump session over time -- a simple metric that will hopefully serve as a useful measure.[^3]

[^3]: Hoping to see a decreasing trend in average heart rate over time indicating an increase in fitness.

```{r}
# Plot of avg. HR data over time.

library(plotly)

jumpGroup <- Hmisc::cut2(jumpData$jumps, cuts = c(501))

p <-
  ggplot(jumpData, aes(start_time, avg_heart_rate, color = jumpGroup,
                       label  = jumps,
                       label2 = jpm,
                       label3 = max_heart_rate,
                       label4 = total_training_effect,
                       label5 = jumpingEffect
                       )
         ) + 
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  # geom_text(vjust = 0, nudge_y = 1.0) +
  labs(title = "Assessing cardio fitness over time from jumping rope",
       subtitle = "Goal is min of 10 min. continuous jumping per session",
       x = "Date",
       y = "Average heart rate (bpm)")

ggplotly(p) %>%
  layout(title = list(text = paste0("Assessing cardio fitness over time from jumping rope",
                                    '<br>',
                                    '<sup>',
                                    "Goal is min of 10 min. continuous jumping per session",
                                    '</sup>')))
```

I've separated the data in the plot into 2 groups: one with points with < 500 jumps and one with >= 500 jumps. This is mainly so the regression line for the >= 500 jump group isn't biased low due to sessions with relatively fewer jumps, which would correspond to a much lower average heart rate.

## Additional exploratory data analysis

### Jumps rate vs. average heart rate by date

```{r}
# Plot of jpm data.

q <-
  ggplot(jumpData, aes(avg_heart_rate, jpm, color = jumpGroup,
                       label  = jumps,
                       label2 = start_time,
                       label3 = max_heart_rate,
                       label4 = total_training_effect,
                       label5 = jumpingEffect
                       )
         ) + 
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  # geom_text(vjust = 0, nudge_y = 1.0) +
  labs(title = "Assessing cardio fitness over time from jumping rope",
       subtitle = "Goal is min of 10 min. continuous jumping per session",
       x = "Average heart rate (bpm)",
       y = "Jumps per minute")

ggplotly(q) %>%
  layout(title = list(text = paste0("Assessing cardio fitness over time from jumping rope",
                                    '<br>',
                                    '<sup>',
                                    "Goal is min of 10 min. continuous jumping per session",
                                    '</sup>')))
```

Viewed a slightly different way, we can highlight the trend using the session date and a color aesthetic. I think this one gets the point across better as the eye can quickly pickup the trend and perceive a clear improvement over time.

```{r}
# Plot of jpm data.

library(gghighlight)
#library(ggrepel)

q <-
  ggplot(jumpData, aes(avg_heart_rate, jpm, color = start_time,
                        # label  = jumps,
                        # label = start_time,
                        # label3 = max_heart_rate,
                        # label4 = total_training_effect
                       # label5 = jumpingEffect
                       )
         ) + 
  geom_point() +
  #geom_smooth(method = lm, se = FALSE) +
  #gghighlight(jpm >= 130) +
  #geom_text(vjust = 0, nudge_y = 1.0) +
  # geom_text_repel(aes(label = as.character(start_time)),
  #                  box.padding   = 0.35,
  #                  point.padding = 0.5,
  #                  segment.color = 'grey50'
  # ) +
  labs(title = "Assessing cardio fitness over time from jumping rope",
       #subtitle = "Highlighted data is jpm >= 130",
       x = "Average heart rate (bpm)",
       y = "Jumps per minute",
       color = "Date")

q

# ggplotly(q) %>%
#   layout(title = list(text = paste0("Change in cardio fitness over time from jumping rope",
#                                     '<br>',
#                                     '<sup>',
#                                     "Goal is min of 10 min. continuous jumping per session",
#                                     '</sup>')))
```

Moving from the bottom right corner of the plot (higher heart rate and lower jump rate) toward the upper left corner (lower heart rate and higher jump rate) shows an increase in jumping efficiency over time (though not necessarily cardio fitness)[^4].

[^4]: This will be visited momentarily looking at recent running data!

### Average heart rate vs. total jumps by date

```{r}
# Plot avg HR vs. number of jumps colored by time.

r <-
  ggplot(jumpData, aes(jumps, avg_heart_rate, color = start_time#,
                       # label  = jpm,
                       # label2 = start_time,
                       # label3 = max_heart_rate,
                       # label4 = total_training_effect,
                       # label5 = jumpingEffect
                       )
         ) + 
  geom_point() +
  gghighlight(jumps >= 1000) +
  geom_smooth(method = lm, se = FALSE) +
  # geom_text(vjust = 0, nudge_y = 1.0) +
  labs(title = "Assessing cardio fitness over time from jumping rope",
       subtitle = "High jump data highlighted",
       x = "Jumps",
       y = "Average heart rate (bpm)",
       color = "Date")

r

# ggplotly(r) %>%
#   layout(title = list(text = paste0("Change in cardio fitness over time from jumping rope",
#                                     '<br>',
#                                     '<sup>',
#                                     "Goal is min of 10 min. continuous jumping per session",
#                                     '</sup>')))
```

Again, the trend is clear by presenting the data this way with the change in color indicating the change in jumping efficiency.

It also shows that:

- The first couple of days were not very productive and
- Jumping inefficiency decreases very fast!

From the very first plot in this article,[^5] it seems that there is about a *10 bpm decrease in average heart rate every 2-3 days in the first week!* This seems to level out as jumping efficiency increases over time leading to a higher jump rate, which causes a slight bump up in average heart rate.

[^5]: It's hard to tell from this plot.

### Using recent hilly training runs to look for cardio gains

Since I've been running while doing this jump rope cardio fitness experiment, there is a good opportunity to use the running data as a basis to look for cardio gains.

Looking at recent running data around the Lalitpur foothills near Kirtipur and Tribhuvan University,[^7] we might be able to detect changes in performance.

[^7]: The runs shown are essentially the same with slight variations due to either time (not enough!) and/or weather (monsoon rains!), which causes a slight (but significant) relative change in total distance.

Using previous plotting tools developed to analyze and compare series' of running data, here is a quick plot of some training runs with some variables of interest.  

```{r}
# Load in all helper functions
source("~/GitHub/myBlog/content/resources/2020-06-14-heat-training-continues-in-hua-hin/RPIandHeatLimitedPacePlots.R")

# Lalitpur Running

dataLalitpur <- getRPIandHLPfromFIT(wd = "~/HealthData/FitFiles/working/", aname = "Lalitpur")
cleanup()

humidity  <-
  tribble(
    ~Date,	~humidity,
    "2020-01-12", 93,
    "2020-08-10", 89,
    "2020-08-11", 100,
    "2020-08-13", 100,
    "2020-08-16", 100,
    "2020-08-18", 100
)

dataLalitpur <- addHI(dataLalitpur, humidity)

plotRPI(dataLalitpur %>% filter(Date > "2020-08-01"), titleText = "Lalitpur Summer 2020 training runs.")
```

A quick glance indicates a high degree of similarity between these runs although it did rain that first day likely providing a significant cooling effect and slight performance bump.[^6]

[^6]: This was also the first day running a longer distance and at higher elevation in nearly 7 months. I did run in Bangkok over the past 5 months, but they were much shorter (~3 miles) due to the intense heat.

Since the running performance index (RPI)[^9] data is so noisy, it is easier to report in tabular form.

[^9]: See [previous post](https://amritpatel.rbind.io/2020/03/27/100-lap-challenge/) for more about how to interpret this running efficiency metric.

```{r, fig.align = 'center'}
library(kableExtra)

dataLalitpur %>% 
  filter(Date > "2020-08-01") %>%
  group_by(Date) %>%
  summarise(AvgRPI = round(mean(RPI, na.rm = T))) %>%
  kable() %>%
  kable_styling(full_width = F, position = "center")
```

Now back to the question:

> *Is jumping rope for 10 minutes a day for 30 days straight improving my cardio fitness?*

Over the same course of time running, we see that the average RPI (or running economy), which is a function of heart rate and pace, is about the same for each run under similar conditions indicating that my performance isn't really improving.

Long story short, this matches what my *Garmin Connect* VO~2~^max^ data indicates since I started running in Kathmandu this summer with a constant value of 48 ml O~2~/kg-min. 

Here is what the *Garmin Connect* app says about VO~2~^max^:

> *VO~2~^max^ is an indication of your cardiovascular fitness and should increase as your level of fitness improves. VO~2~^max^ is the maximum volume of oxygen (in milliliters) you can consume per minute per kilogram of body weight at your maximum performance. Your compatible Garmin device uses heart rate and exercise data to estimate your VO~2~^max^.*

And what is really great (from the [*Forerunner 935 Owner's Manual*](https://www8.garmin.com/manuals/webhelp/forerunner935/EN-US/GUID-F5BF67CE-C94E-4842-AE96-A7A05C85B732.html)):

> *Your device adjusts the VO~2~^max^ values for heat and altitude, such as when you are acclimating to high heat environments or high altitude.*

So accounting for differences in elevation gives added confidence in the *Garmin*-calculated VO~2~^max^.

## tldr

Jumping rope is great for the reasons listed above *except* it didn't really improve my cardio fitness.[^10] Maybe this is because I already had high cardio fitness? Maybe it would be different if I didn't?

[^10]: I guess I could have just looked at the Garmin "training effect" data for the jump rope sessions to see that the "total training effect" values are practically 0! Yet another point of training effect validation...

```{r, eval = F}
plotViolationSumm(dataLalitpur %>% filter(Date > "2020-08-01"), titleText = "Lalitpur Summer 2020 training runs.")
```

## Notes

### 2020-08-12

I ran 6.72 miles in morning; average heart rate was 147 bpm.

### 2020-08-13

While jumping rope today, I naturally started to move my feet differently. I began alternating my feet instead of jumping both at the same time early on in the 10 minute session. This resulted in much less of a bounce height. This also led to much more consistency while jumping with minimal disruptions of jump streaks. ***I had the realization that the average heart rate data reflected in the plot above is likely more indicative of my body learning to jump more efficiently rather than my cardio fitness improving.***

I'm also wondering what the impact of me running relatively long distances recently has on the jump rope data. Does this introduce an element of fatigue or hysteresis that causes my average heart rate data on running days to be biased high when it otherwise wouldn't be? Needless to say, the first day I went running (yesterday) my calves felt really tight (due to jumping rope every day for a week). However, I noticed that in the morning today, they actually felt better (less tight and sore), as if running helped loosen them up. This is counter to what I would expect though because typically running causes me to have tighter calves. I looked to the internet to try and explain what was going on and found an [article](https://blog.mapmyrun.com/5-simple-solutions-for-tight-calves/#:~:text=A%20good%20way%20to%20stretch,faster%2C%20in%20a%20pumping%20motion.) mentioning that:

> *A good way to stretch your calves is during your workout: Walking or running uphill is a great calf activator and naturally forces your muscles to stretch while your heading up.*

Since the recent running route is about half uphill, I might be getting some benefit from this.

### 2020-08-14

I ran 5.88 miles in morning; average heart rate was 149 bpm.

### 2020-08-18

Need to account for temperature effects.

### 2020-08-19

Jumped in rain, which served as a cooling effect; temperature conditions similar to yesterday.

### 2020-08-20

Fever of 100.6-100.9 F measured around 8 pm. Temperature recorded as normal (98.5 F) when woke up next morning around 7 am.